Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-11 paper_io_sim.tex
prettified tables

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{graphicx}
\usepackage{paralist}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage[dvipsnames]{xcolor}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{tabularx}

@@

    % Henri Rewrote quite  a bit in the next paragraph here to spell out a bit
    % more why we need simulation
    The emergence of Big Data in recent years has resulted in a growing
    need for efficient data storage management techniques.  While
    infrastructures with sufficient compute power are available to
    process large datasets, the I/O bottleneck remains.  A
    well-known and ubiquitous approach for reducing the overhead of
    I/O operations is to use the Linux page cache.  It is thus crucial
    to study experimentally the impact of the Linux page cache on the
    performance Big Data applications. Given the difficulties with and shortcomings of
    real-world experiments, a popular approach for application
    performance studies is to use simulation.  While many simulation
    frameworks have been developed for this purpose, they do not 
        simulate page caching fully, or even at all.  As a result, using
    these frameworks for studying the performance of data-intensive
    applications leads inaccurate simulation results.

        % Some cosmetic rewrites below
    In this paper, we propose a I/O simulation model that characterizes
    the key features of the page cache. We have implemented this model
    as  part of the \wrench workflow simulation framework, which itself
    builds on the popular \simgrid distributed systems simulation
    framework.  Our model and its implementation enable the simulation
    of both single-threaded and multithreaded applications, and of both
    writeback and writethrough caches for local or network-based
    filesystems. We evaluate the accuracy of our model in different
    conditions, including sequential and concurrent applications, as
    well as local and remote I/Os.  We find that our page cache model
    reduces the simulation error by up to an order of magnitude when
    compared to state-of-the-art, cacheless simulations.

@@

        % Swapped the order the two paragraphs and merged them into one
        The number of data-intensive applications has been steadily rising as a result of
        open-data and data sharing initiatives. Due to the sheer size of the data being
        processed, these applications must be executed on large scale infrastructure
    such as High Performance Computing (HPC) clusters or the cloud.  It
    is thus crucial to quantify the performance of these applications
    on these platforms. The goals include determining which type of hardware/software
    stacks are best suited to different application classes, as well as
    understanding the shortcomings of current algorithms, designs and
    technologies so as to propose new solutions. Unfortunately, application performance
        studies via experiments on real-world compute platforms
        faces several difficulties (high operational costs, labor-intensive experimental setups,
        shared platforms with dynamic loads that hinder reproducibility of results) and shortcomings
        (experiments are limited to the available platform/software configurations, which precludes
        the exploration of hypothetical scenarios). 
    Simulations address these concerns by providing simulation models
    and abstractions for the performance of computer hardware, such as
    CPU, network and storage. Consequently, simulations provide a
    cost-effective, fast, easy and reproducible way to evaluate
    application performance on arbitrary platform configurations. It thus comes
        as no surprise that, considering here only the field of parallel and distributed
        computing, a large number of simulation frameworks have been developed and used
        to support research and development endeavors~\cite{ optorsim, gridsim, groudsim, cloudsim, nunez2012simcan,nunez2012icancloud, mdcsim, dissect_cf, cloudnetsimplusplus, fognetsimplusplus, casanova2014simgrid, ROSS, casanova2020fgcs}.

        Page caching is a well-known and ubiquitous technique for mitigating the I/O bottleneck
        faced by data-intensive applications. As such, it is necessary to simulate page caching
        when performing performance studies in simulation.
        Furthermore, simulating the page cache makes it possible to study relevant application-I/O
        bottlenecks that can arise over the course of execution, such as page cache
        starvation~\cite{zhuang2017}.
        While existing simulation frameworks of parallel and distributed computing
        systems  capture many relevant features of hardware/software stacks, 
        they lack the ability to simulate page caches adequately. Those frameworks
        that implement some page cache simulation
        fail to capture key features such
        as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
        Some simulators, such as that in~\cite{xu2018saving}, do capture such features
        but are domain-specific.

@@

        To improve performance of disk accesses, the page cache was introduced to the Linux kernel. %It contains pages referring to physical pages on disk \cite{linuxdev3rd2010}.
        The goal of the page cache is to try ensure that most I/O operations occur entirely
        within memory space rather than synchronously to and from the storage device.
        All disk-backed data located in the page cache are partitioned into many small pages.
        When a read operation is initiated, the kernel checks whether the required pages are in memory.
        If it is the case, data is read entirely from memory (i.e. \textit{cache hit}). Otherwise,
        the necessary pages must be loaded to memory from the disk (i.e. \textit{cache miss})
        Conversely, for writing, one of three strategies can be applied: direct~I/O (or, ``no-write''),
        writethrough and writeback.
        If direct~I/O is enabled for a given file system, all I/O operations bypass the
        page cache for writes, writing data directly to disk. In contrast, writethrough enables the page cache
        for file system writes, but the data is written synchronously from memory to disk. Writeback extends writethrough
        by enabling asynchronous writes to disk, meaning that a write call will return once
        the data has been written to memory.
        The writeback strategy is considered to outperform writethrough as well as
        direct I/O as it delays disk writes to perform a bulk write at a later time
        \cite{linuxdev3rd2010}.
        Cache eviction and flushing strategies are integral to proper page cache functioning.
        Whenever space in memory becomes limited, either as a result of application memory
        or page cache use, page cache data may be evicted. Only data that
        has been persisted to storage (clean pages) can be flagged for eviction and removed from
        memory. Written data that has not yet been persisted to disk (dirty data) must first
        be copied (flushed) to storage prior to eviction. When sufficient memory is
        being occupied, the flushing process is synchronous. However, even when
        there is sufficient available memory, written data will be flushed to disk
        at a predefined interval through a process known as \textit{periodical flushing}.
        Periodical flushing only flushes expired dirty pages, which remain dirty in
        page cache longer than an expiration time configured in the kernel.
        Different cache eviction algorithms have also been developed and proposed
        \cite{owda2014comparison}.
        The Linux kernel uses a two-list strategy to flag pages for eviction.
        The two-list strategy is based on a least recently used (LRU) policy
        and uses an active and inactive list in its implementation.
        If accessed pages are not in the page cache, they are added to the inactive list.
        Should pages located on the inactive list be accessed, they will be moved from
        the inactive to the active list.
        The lists are also kept balanced by moving pages from the active list
        to the inactive list when the active list grows too large.
        Thus, the active list only contains pages which are accessed more than once
        and not evictable, while the inactive list includes pages accessed once only,
        or pages that have be accessed more than once but moved from the active list.
        Both lists operate using LRU eviction policies meaning the data that has
        not be accessed recently will be moved first.

@@

        The two main concerns for simulation are \emph{accuracy}
        (the ability to faithfully reproduce real-world executions) and
        \emph{scalability}  (the ability to simulate large/long real-world
        executions quickly and with low RAM footprint), and the above
        frameworks achieve different compromises between the two.  At
        one extreme are discrete-event models that capture
        ``microscopic'' behaviors of hardware/software systems (e.g.,
        packet-level network simulation, block-level disk simulation,
        cycle-accurate CPU simulation), which favor accuracy over
        speed.  At the other extreme are analytical models that capture
        ``macroscopic'' behaviors via mathematical models.  While these
        models lead to fast simulation, they must be developed
        carefully if high levels of accuracy are to be
        achieved~\cite{velhoTOMACS2013}.

@@

        To parametrize our simulators, we benchmarked the
        memory, local disk, remote disk (NFS), and network bandwidths
        (Table~\ref{table:benchmark}). Since \simgrid, and thus \wrench, currently only supports
        symmetrical bandwidths, we use the mean of the read and write
        bandwidth values in our experiments.
            \begin{table}[htbp]
            \centering
            \begin{tabularx}{\columnwidth}{|l
            |>{\centering\arraybackslash}X
            |>{\centering\arraybackslash}X
            |>{\centering\arraybackslash}X|}
            \hline
                Bandwidths  & Cluster (real) & Python simulator & \wrench simulator\\
            \hline
                Memory read  & 6860    & 4812     & 4812\\
                Memory write & 2764    & 4812 & 4812\\
                Local disk read & 510 & 465 & 465\\
                Local disk write & 420 & 465     & 465\\
                Remote disk read & 515 & - & 445\\
                Remote disk write & 375 & - & 445\\
                Network & 3000 & - & 3000\\
            \hline
            \end{tabularx}
            \caption{Bandwidth benchmarks (MBps) and simulator configurations.
            The bandwidths used in the simulations are the average of the read and write bandwidths
            measured in the cluster.
            Remote disk and network accesses are not simulated in the Python simulator. \tristan{table could be prettified}}
            \label{table:benchmark}
            \end{table}

@@

            \begin{table}[htbp]
            \centering
            \begin{tabularx}{0.8\columnwidth}{|l|>{\centering\arraybackslash}X|}
            \hline
                Input size (GB)  & CPU time (second)\\
            \hline
                3      & 4.4 \\
                20  & 28 \\
                50  & 75 \\
                75  & 110 \\
                100  & 155 \\
            \hline
            \end{tabularx}
            \caption{Task CPU time corresponding to input file size. \tristan{are the input files in GB or in GiB? If GiB, then change the abbreviation everywhere.}
            \tristan{add nighres CPU times when available}}
            \label{table:cputime}
            \end{table}

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{graphicx}
\usepackage{paralist}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage[dvipsnames]{xcolor}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{multirow}

@@

    % Henri Rewrote quite  a bit in the next paragraph here to spell out a bit
    % more why we need simulation
    The emergence of Big Data in recent years has resulted in a growing
    need for efficient data storage management techniques.  While
    infrastructures with sufficient compute power are available to
    process large datasets, the I/O bottleneck remains.  A
    well-known approach for reducing the overhead of
    I/O operations is to use the Linux page cache.  It is thus crucial
    to study experimentally the impact of the Linux page cache on the
    performance of Big Data applications. Given the difficulties with and shortcomings of
    real-world experiments, a popular approach for application
    performance studies is to use simulation.  While many simulation
    frameworks have been developed for this purpose, they do not 
    simulate page caching fully, or even at all.  As a result, using
    these frameworks for studying the performance of data-intensive
    applications leads to inaccurate simulation results.

    In this paper, we propose a I/O simulation model that characterizes
    the key features of the page cache. We have implemented this model
    as  part of the \wrench workflow simulation framework, which itself
    builds on the popular \simgrid distributed systems simulation
    framework.  Our model and its implementation enable the simulation
    of both single-threaded and multithreaded applications, and of both
    writeback and writethrough caches for local or network-based
    filesystems. We evaluate the accuracy of our model in different
    conditions, including sequential and concurrent applications, as
    well as local and remote I/Os.  We find that our page cache model
    reduces the simulation error by up to an order of magnitude when
    compared to state-of-the-art, cacheless simulations.

@@

        % Swapped the order the two paragraphs and merged them into one
        The number of data-intensive applications has been steadily rising as a result of
        open-data and data sharing initiatives. Due to the sheer size of the data being
        processed, these applications must be executed on large-scale infrastructures
    such as High Performance Computing (HPC) clusters or the cloud.  It
    is thus crucial to quantify the performance of these applications
    on these platforms. The goals include determining which type of hardware/software
    stacks are best suited to different application classes, as well as
    understanding the shortcomings of current algorithms, designs and
    technologies so as to propose new solutions. Unfortunately, application performance
        studies via experiments on real-world compute platforms
        face several difficulties (high operational costs, labor-intensive experimental setups,
        shared platforms with dynamic loads that hinder reproducibility of results) and shortcomings
        (experiments are limited to the available platform/software configurations, which precludes
        the exploration of hypothetical scenarios). 
    Simulations address these concerns by providing models
    and abstractions for the performance of computer hardware, such as
    CPU, network and storage. As a result, simulations provide a
    cost-effective, fast, easy and reproducible way to evaluate
    application performance on arbitrary platform configurations. It thus comes
        as no surprise that, considering here only the field of parallel and distributed
        computing, a large number of simulation frameworks have been developed and used
        to support research and development endeavors~\cite{ optorsim, gridsim, groudsim, cloudsim, nunez2012simcan,nunez2012icancloud, mdcsim, dissect_cf, cloudnetsimplusplus, fognetsimplusplus, casanova2014simgrid, ROSS, casanova2020fgcs}.

        Page caching is an ubiquitous technique for mitigating the I/O bottleneck
        faced by data-intensive applications. As such, it is necessary to simulate page caching
        when performing performance studies in simulation.
        Furthermore, simulating the page cache makes it possible to study relevant application-I/O
        bottlenecks that can arise over the course of execution, such as page cache
        starvation~\cite{zhuang2017}.
        While existing simulation frameworks of parallel and distributed computing
        systems  capture many relevant features of hardware/software stacks, 
        they lack the ability to simulate page cache adequately. Those frameworks
        that implement some page cache simulation
        fail to capture key features such
        as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
        Some simulators, such as that in~\cite{xu2018saving}, do capture such features
        but are domain-specific.

@@

        To improve performance of disk accesses, the page cache was introduced to the Linux kernel. %It contains pages referring to physical pages on disk \cite{linuxdev3rd2010}.
        The goal of the page cache is to ensure that most I/O operations occur entirely
        within memory space rather than synchronously to and from storage devices.
        All disk-backed data located in the page cache are partitioned into many small pages.
        When a read operation is initiated, the kernel checks whether the required pages are in memory.
        If it is the case, data is read entirely from memory (i.e. \textit{cache hit}). Otherwise,
        the necessary pages must be loaded to memory from the disk (i.e. \textit{cache miss}).
        Conversely, for writing, one of three strategies can be applied: direct~I/O (or, ``no-write''),
        writethrough and writeback.
        If direct~I/O is enabled for a given file system, all I/O operations bypass the
        page cache for writes, writing data directly to disk. In contrast, writethrough enables the page cache
        for file system writes, but the data is written synchronously from memory to disk. Writeback extends writethrough
        by enabling asynchronous writes to disk, meaning that a write call will return once
        the data has been written to memory.
        The writeback strategy is considered to outperform writethrough as well as
        direct I/O as it delays disk writes to perform a bulk write at a later time
        \cite{linuxdev3rd2010}.
        Cache eviction and flushing strategies are integral to proper page cache functioning.
        Whenever space in memory becomes limited, either as a result of application memory
        or page cache use, page cache data may be evicted. Only data that
        has been persisted to storage (clean pages) can be flagged for eviction and removed from
        memory. Written data that has not yet been persisted to disk (dirty data) must first
        be copied (flushed) to storage prior to eviction. When sufficient memory is
        being occupied, the flushing process is synchronous. However, even when
        there is sufficient available memory, written data will be flushed to disk
        at a predefined interval through a process known as \textit{periodical flushing}.
        Periodical flushing only flushes expired dirty pages, which remain dirty in
        page cache longer than an expiration time configured in the kernel.
        Different cache eviction algorithms have also been developed and proposed
        \cite{owda2014comparison}.
        The Linux kernel uses a two-list strategy to flag pages for eviction.
        The two-list strategy is based on a least recently used (LRU) policy
        and uses an active and inactive list in its implementation.
        If accessed pages are not in the page cache, they are added to the inactive list.
        Should pages located on the inactive list be accessed, they will be moved from
        the inactive to the active list.
        The lists are also kept balanced by moving pages from the active list
        to the inactive list when the active list grows too large.
        Thus, the active list only contains pages which are accessed more than once
        and not evictable, while the inactive list includes pages accessed once only,
        or pages that have be accessed more than once but moved from the active list.
        Both lists operate using LRU eviction policies meaning the data that has
        not be accessed recently will be moved first.

@@

        The two main concerns for simulation are \emph{accuracy}
        (the ability to faithfully reproduce real-world executions) and
        \emph{scalability}  (the ability to simulate large/long real-world
        executions quickly and with low RAM footprint). The above
        frameworks achieve different compromises between the two.  At
        one extreme are discrete-event models that capture
        ``microscopic'' behaviors of hardware/software systems (e.g.,
        packet-level network simulation, block-level disk simulation,
        cycle-accurate CPU simulation), which favor accuracy over
        speed.  At the other extreme are analytical models that capture
        ``macroscopic'' behaviors via mathematical models.  While these
        models lead to fast simulation, they must be developed
        carefully if high levels of accuracy are to be
        achieved~\cite{velhoTOMACS2013}.

@@

        To parametrize our simulators, we benchmarked the
        memory, local disk, remote disk (NFS), and network bandwidths
        (Table~\ref{table:benchmark}). Since \simgrid, and thus \wrench, currently only supports
        symmetrical bandwidths, we use the mean of the read and write
        bandwidth values in our experiments.
            \begin{table}
            \centering
            \begin{tabularx}{\columnwidth}{ll
            >{\centering\arraybackslash}X
            >{\centering\arraybackslash}X
            >{\centering\arraybackslash}X}
            \toprule
                \multicolumn{2}{c}{Bandwidths}  & Cluster (real) & Python prototype & \wrench simulator\\
            \midrule
            \multirow{2}{*}{Memory}      & read  & 6860 & 4812 & 4812\\
                                         & write & 2764 & 4812 & 4812\\
            \multirow{2}{*}{Local disk}  & read  & 510  & 465  & 465\\
                                         & write & 420  & 465  & 465\\
            \multirow{2}{*}{Remote disk} & read  & 515  & -    & 445\\
                                         & write & 375  & -    & 445\\
            \multicolumn{2}{l}{Network}  & 3000  & -    & 3000\\
            \bottomrule
            \end{tabularx}
            \caption{Bandwidth benchmarks (MBps) and simulator configurations.
            The bandwidths used in the simulations are the average of the read and write bandwidths
            measured in the cluster.
            Remote disk and network accesses are not simulated in the Python simulator.}
            \label{table:benchmark}
            \end{table}

@@

            \begin{table}[b]
            \centering
            \begin{tabularx}{0.8\columnwidth}{c>{\centering\arraybackslash}X}
            \toprule
                Input size (GB)  & CPU time (second)\\
            \midrule
                3      & 4.4 \\
                20  & 28 \\
                50  & 75 \\
                75  & 110 \\
                100  & 155 \\
            \bottomrule
            \end{tabularx}
            \caption{Task CPU times used in the simulation. \tristan{are the input files in GB or in GiB? If GiB, then change the abbreviation everywhere.}
            \tristan{add nighres CPU times when available}}
            \label{table:cputime}
            \end{table}
